\documentclass{ltxdoc}
% \usepackage{tgschola,url}
\usepackage{url}
\usepackage[english]{babel}
\usepackage{hyperref}
\usepackage{luacode}
\usepackage{framed}
% Version is defined in the makefile, use default values when compiled directly
\ifdefined\version\else
\def\version{0}
\let\gitdate\date
\fi
\newcommand\modulename[1]{\subsection{#1}\label{sec:#1}}
\newcommand\modulesummary[1]{#1\\}
\newcommand\moduleclass[1]{\subsubsection{Class: #1}}
\newcommand\functionname[2]{\par\noindent\textbf{#1(#2)}\\}
\newcommand\functionsummary[1]{#1\\\textbf{Parameters:}\\}
\newcommand\functionparam[2]{\texttt{#1}: #2\\}
\newcommand\functionreturn[1]{\textbf{Return: }\\#1\\}

\begin{document}
	\title{The \textsc{LuaXML} library}
	\author{Paul Chakravarti \and Michal Hoftich}
	\date{Version \version\\\gitdate}
	\maketitle
\tableofcontents

\section{Introduction}

|LuaXML| is pure lua library for processing and serializing of the |xml| files.
The base code code has been written by Paul Chakravarti, with minor changes
which brings Lua 5.3 or HTML 5 support. On top of that, new modules for
accessing the |xml| files using |DOM| like methods or |CSS|
selectors\footnote{Thanks to Leaf Corcoran for |CSS selector| parsing code.}
have been added.

The documentation is divided to three parts -- first part deals with the |DOM|
library, second part describes the low-level libraries and the third part is
original documentation by Paul Chakravarti.
% Current release is aimed mainly as support for the odsfile package. 
% In first release it was included with the odsfile package,
% but as it is general library which can be used also with other packages, 
% I decided to distribute it as separate library.

\section{The \texttt{DOM\_Object} library}

This library can process a |xml| sources using |DOM| like functions. To load
it, you need to require |luaxml-domobject.lua| file. The |parse| function
provided by the library creates \texttt{DOM\_Object} object, which provides several
methods for processing the |xml| tree.

\begin{verbatim}
local dom = require "luaxml-domobject"
local document = [[
<html>
<head><title>sample</title></head>
<body>
<h1>test</h1>
<p>hello</p>
</body>
</html>
  ]]

-- dom.parse returns the DOM_Object
local obj = dom.parse(document)
-- it is possible to call methods on the object
local root_node = obj:root_node()
for _, x in ipairs(root_node:get_children()) do
  print(x:get_element_name())
end
\end{verbatim}

The details about available methods can be found in the API docs, section
\ref{sec:luaxml-domobject}. The above code will load a |xml| document, it will
get the ROOT element and print all it's children element names.  The
\verb|DOM_Object:get_children| function returns Lua table, so it is possible to
loop over it using standard table functions.

\begin{framed}
\begin{luacode*}
dom = require "luaxml-domobject"
local document = [[
<html>
<head><title>sample</title></head>
<body>
<h1>test</h1>
<p>hello</p>
</body>
</html>
  ]]

-- dom.parse returns the DOM_Object
obj = dom.parse(document)
-- it is possible to call methods on the object
local root_node = obj:root_node()
for _, x in ipairs(root_node:get_children()) do
  tex.print(x:get_element_name().. "\\par")
end
\end{luacode*}
\end{framed}

\subsection{Node selection methods}
There are some other methods for element retrieving. 

\subsubsection{The \texttt{DOM\_Object:get\_path} method}
If you want to print text content of all child elements of the body element, you can use \verb|DOM_Object:get_path|:

\begin{verbatim}
local path = obj:get_path("html body")
for _, el in ipairs(path[1]:get_children()) do
  print(el:get_text())
end
\end{verbatim}

The \verb|DOM_Object:get_path| function always return array with all elements
which match the requested path, even it there is only one such element. In this
case, it is possible to use standard Lua table indexing to get the first and
only one matched element and get it's children using
\verb|DOM_Object:get_children| method. It the children node is an element, it's
text content is printed using \verb|DOM_Object:get_text|.



\begin{framed}
  \begin{luacode*}
local path = obj:get_path("html body")

for _, el in ipairs(path[1]:get_children()) do
  if el:is_element() then
    tex.print(el:get_text().."\\par")
  end
end
  \end{luacode*}
\end{framed}

\subsubsection{The \texttt{DOM\_Object:query\_selector} method}

This method uses |CSS selector| syntax to select elements, similarly to JavaScript \textit{jQuery} library.

\begin{verbatim}
for _, el in ipairs(obj:query_selector("h1,p")) do
  print(el:get_text())
end
\end{verbatim}


\begin{framed}
  \begin{luacode*}
for _, el in ipairs(obj:query_selector("h1,p")) do
  tex.print(el:get_text().."\\par")
end
  \end{luacode*}
\end{framed}

\subsection{Element traversing}

\subsubsection{The \texttt{DOM\_Object:traverse\_elements} method}

It may be useful to traverse over all elements and apply a function on all of them. 

\begin{verbatim}
obj:traverse_elements(function(node)
  print(node:get_text())
end)
\end{verbatim}

\begin{framed}
  \begin{luacode*}
obj:traverse_elements(function(node)
  tex.print(node:get_text().."\\par")
end)
  \end{luacode*}
\end{framed}

The \verb|get_text| method gets text from all children elements, so the first
line shows all text contained in the \verb|<html>| element, the second one in
\verb|<head>| element and so on.  

\subsection{DOM modifications}

It is possible to add new elements, text nodes, or to remove them. 

\begin{verbatim}
local headers = obj:query_selector("h1")
for _, header in ipairs(headers) do
  header:remove_node()
end
-- query selector returns array, we must retrieve the first element
-- to get the actual body element
local body = obj:query_selector("body")[1]
local paragraph = body:create_element("p", {})
body:add_child_node(paragraph)
paragraph:add_child_node(paragraph:create_text_node("This is a second paragraph"))

for _, el in ipairs(body:get_children()) do
  if el:is_element() then
    print(el:get_element_name().. ": ".. el:get_text())
  end
end
\end{verbatim}

In this example, \verb|<h1>| element is being removed from the sample document, and new 
paragraph is added. Two paragraphs should be shown in the output:

\begin{framed}
  \begin{luacode*}
local headers = obj:query_selector("h1")
-- query selector returns array, we must retrieve the first element
-- to get the actual body element
local body = obj:query_selector("body")[1]
local oldbody = body:copy_node()
for _, header in ipairs(headers) do
  header:remove_node()
end
local paragraph = body:create_element("p", {})
body:add_child_node(paragraph)
paragraph:add_child_node(paragraph:create_text_node("This is a second paragraph"))

for _, el in ipairs(body:get_children()) do
if el:is_element() then
  tex.print(el:get_element_name().. ": ".. el:get_text() .. "\\par")
end
end

body:replace_node(oldbody)
  \end{luacode*}
\end{framed}


\section{The API documentation}

\input{doc/api.tex}

\section{Low-level functions usage}

% The processing is done with several handlers, their usage will be shown in the
% following section. Full description of handlers is given in the original
% documentation in section \ref{sec:handlers}.

% \subsection{Usage examples}

The original |LuaXML| library provides some low-level functions for |XML| handling.
First of all, we need to load the libraries:

\begin{verbatim}
xml = require('luaxml-mod-xml')
handler = require('luaxml-mod-handler')
\end{verbatim} 


The |luaxml-mod-xml| file contains the  xml parser and also the serializer. In
|luaxml-mod-handler|, various handlers for dealing with xml data are defined.
Handlers transforms the |xml| file to data structures which can be handled from
the Lua code. More information about handlers can be found in the original
documentation, section \ref{sec:handlers}.

\subsection{The simpleTreeHandler} 
\begin{verbatim}
sample = [[
<a>
  <d>hello</d>
  <b>world.</b>
  <b at="Hi">another</b>
</a>]]
treehandler = handler.simpleTreeHandler()
x = xml.xmlParser(treehandler)
x:parse(sample)
\end{verbatim} 

You have to create handler object, using |handler.simpleTreeHandler()| and xml
parser object using |xml.xmlParser(handler object)|. |simpleTreehandler|
creates simple table hierarchy, with top root node in |treehandler.root|

\begin{verbatim}
-- pretty printing function
function printable(tb, level)
  level = level or 1
  local spaces = string.rep(' ', level*2)
  for k,v in pairs(tb) do
    if type(v) ~= "table" then
      print(spaces .. k..'='..v)
    else
      print(spaces .. k)
      level = level + 1
      printable(v, level)
    end
  end
end

-- print table
printable(treehandler.root)
-- print xml serialization of table
print(xml.serialize(treehandler.root))
-- direct access to the element
print(treehandler.root["a"]["b"][1])
\end{verbatim}

This code produces the following output:

\begin{verbatim}
 output:
   a
     d=hello
     b
       1=world.
       2
         1=another
         _attr
           at=Hi
 <?xml version="1.0" encoding="UTF-8"?>
 <a>
   <d>hello</d>
     <b>world.</b>
     <b at="Hi">
       another
     </b>
 </a>
 
 world.
\end{verbatim}

First part is pretty-printed dump of Lua table structure contained in the handler, the second
part is |xml| serialized from that table and the last part demonstrates direct access to particular
elements.

Note that |simpleTreeHandler| creates tables that can be easily accessed using
standard lua functions, but if the xml document is of mixed-content type\footnote{%
This means that element may contain both children elements and text.}:

\begin{verbatim}
<a>hello
  <b>world</b>
</a>	  
\end{verbatim}

\noindent then it produces wrong results. It is useful mostly for data |xml| files, not for
text formats like |xhtml|.

\subsection{The domHandler}

% For complex xml documents with mixed content, |domHandler| is capable of representing any valid XML document:
For complex xml documents, it is best to use the |domHandler|, which creates object which contains all information
from the |xml| document. 

\begin{verbatim}
-- file dom-sample.lua
-- next line enables scripts called with texlua to use luatex libraries
--kpse.set_program_name("luatex")
function traverseDom(current,level)
  local level = level or 0
  local spaces = string.rep(" ",level)
  local root= current or current.root
  local name = root._name or "unnamed"
  local xtype = root._type or "untyped"
  local attributes = root._attr  or {} 
  if xtype == "TEXT" then 
    print(spaces .."TEXT : " .. root._text)
  else	 
    print(spaces .. xtype .. " : " .. name) 
  end
  for k, v in pairs(attributes) do
    print(spaces .. "  ".. k.."="..v)
  end
  local children = root._children or {}
  for _, child in ipairs(children) do
    traverseDom(child, level + 1)
  end
end

local xml = require('luaxml-mod-xml')
local handler = require('luaxml-mod-handler')
local x = '<p>hello <a href="http://world.com/">world</a>, how are you?</p>'
local domHandler = handler.domHandler()
local parser = xml.xmlParser(domHandler)
parser:parse(x)
traverseDom(domHandler.root)
\end{verbatim}

The ROOT element is stored in |domHandler.root| table, it's child nodes are stored in |_children|
tables. Node type is saved in |_type| field, if the node type is |ELEMENT|, then |_name| field contains 
element name, |_attr| table contains element attributes. |TEXT| node contains text content in |_text| 
field.

The previous code produces following output in the terminal: % after command
% |texlua dom-sample.lua| running:

\begin{verbatim}
ROOT : unnamed
 ELEMENT : p
  TEXT : hello
  ELEMENT : a
    href=http://world.com/
   TEXT : world
  TEXT : , how are you?
\end{verbatim}

% With \verb|domHandler|, you can process documents with mixed content, like
% \verb|xhtml|, so it is a most powerful handler.


\clearpage
\part{Original \texttt{LuaXML} documentation by Paul Chakravarti}
\medskip

\noindent This document was created automatically from the original source code comments using Pandoc\footnote{\url{http://johnmacfarlane.net/pandoc/}} 

\section{Overview}


This module provides a non-validating XML stream parser in Lua. 
\section{Features}

\begin{itemize}
\item
  Tokenises well-formed XML (relatively robustly)
\item
  Flexible handler based event api (see below)
\item
  Parses all XML Infoset elements - ie.
  \begin{itemize}
  \item
    Tags
  \item
    Text
  \item
    Comments
  \item
    CDATA
  \item
    XML Decl
  \item
    Processing Instructions
  \item
    DOCTYPE declarations
  \end{itemize}
\item
  Provides limited well-formedness checking (checks for basic syntax \&
  balanced tags only)
\item
  Flexible whitespace handling (selectable)
\item
  Entity Handling (selectable)
\end{itemize}
\section{Limitations}

\begin{itemize}
\item
  Non-validating
\item
  No charset handling
\item
  No namespace support
\item
  Shallow well-formedness checking only (fails to detect most semantic
  errors)
\end{itemize}
\section{API}

The parser provides a partially object-oriented API with functionality
split into tokeniser and hanlder components.

The handler instance is passed to the tokeniser and receives callbacks
for each XML element processed (if a suitable handler function is
defined). The API is conceptually similar to the SAX API but implemented
differently.

The following events are generated by the tokeniser

\begin{verbatim}
handler:starttag       - Start Tag
handler:endtag         - End Tag
handler:text        - Text
handler:decl        - XML Declaration
handler:pi          - Processing Instruction
handler:comment     - Comment
handler:dtd         - DOCTYPE definition
handler:cdata       - CDATA 
\end{verbatim}
The function prototype for all the callback functions is

\begin{verbatim}
callback(val,attrs,start,end)
\end{verbatim}
where attrs is a table and val/attrs are overloaded for specific
callbacks - ie.

\begin{tabular}{llp{5cm}}
Callback   &  val        &    attrs (table)\\
\hline
starttag     &   name &   |{ attributes (name=val).. }|\\
endtag       &   name   &    nil\\
text      &   |<text>| &   nil\\
cdata     &   |<text> |  &   nil\\
decl      &   "xml"       &   |{ attributes (name=val).. }|\\
pi        &   pi name     &  \begin{verbatim}{ attributes (if present)..
  _text = <PI Text>
}\end{verbatim}\\
comment   &   |<text>|      &   nil\\     
dtd       &   root element  & \begin{verbatim}{ _root = <Root Element>,
  _type = SYSTEM|PUBLIC,
  _name = <name>,
  _uri = <uri>,
  _internal = <internal dtd>
}\end{verbatim}\\
\end{tabular}

(starttag \& endtag provide the character positions of the start/end of the
element)

XML data is passed to the parser instance through the `parse' method
(Note: must be passed as single string currently)

\section{Options}

Parser options are controlled through the `self.options' table.
Available options are -

\begin{itemize}
\item
  stripWS

  Strip non-significant whitespace (leading/trailing) and do not
  generate events for empty text elements
\item
  expandEntities

  Expand entities (standard entities + single char numeric entities only
  currently - could be extended at runtime if suitable DTD parser added
  elements to table (see obj.\_ENTITIES). May also be possible to expand
  multibyre entities for UTF--8 only
\item
  errorHandler

  Custom error handler function
\end{itemize}
NOTE: Boolean options must be set to `nil' not `0'

\section{Usage}

Create a handler instance -

\begin{verbatim}
h = { starttag = function(t,a,s,e) .... end,
      endtag = function(t,a,s,e) .... end,
      text = function(t,a,s,e) .... end,
      cdata = text }
\end{verbatim}
(or use predefined handler - see luaxml-mod-handler.lua)

Create parser instance -

\begin{verbatim}
p = xmlParser(h)
\end{verbatim}
Set options -

\begin{verbatim}
p.options.xxxx = nil
\end{verbatim}
Parse XML data -

\begin{verbatim}
xmlParser:parse("<?xml... ")
\end{verbatim}
\section{Handlers}\label{sec:handlers}

\subsection{Overview}

Standard XML event handler(s) for XML parser module (luaxml-mod-xml.lua)

\subsection{Features}

\begin{verbatim}
printHandler        - Generate XML event trace
domHandler          - Generate DOM-like node tree
simpleTreeHandler   - Generate 'simple' node tree
simpleTeXhandler    - SAX like handler with support for CSS selectros
\end{verbatim}
\subsection{API}

Must be called as handler function from xmlParser and implement XML
event callbacks (see xmlParser.lua for callback API definition)

\subsubsection{printHandler}

printHandler prints event trace for debugging

\subsubsection{domHandler}

domHandler generates a DOM-like node tree  structure with 
a single ROOT node parent - each node is a table comprising 
fields below.

\begin{verbatim}
node = { _name = <Element Name>,
        _type = ROOT|ELEMENT|TEXT|COMMENT|PI|DECL|DTD,
        _attr = { Node attributes - see callback API },
        _parent = <Parent Node>
        _children = { List of child nodes - ROOT/NODE only }
      }

\end{verbatim}
\subsubsection{simpleTreeHandler}

simpleTreeHandler is a simplified handler which attempts to generate a
more `natural' table based structure which supports many common XML
formats.

The XML tree structure is mapped directly into a recursive table
structure with node names as keys and child elements as either a table
of values or directly as a string value for text. Where there is only a
single child element this is inserted as a named key - if there are
multiple elements these are inserted as a vector (in some cases it may
be preferable to always insert elements as a vector which can be
specified on a per element basis in the options). Attributes are
inserted as a child element with a key of `\_attr'.

Only Tag/Text \& CDATA elements are processed - all others are ignored.

This format has some limitations - primarily

\begin{itemize}
\item  Mixed-Content behaves unpredictably - the relationship between text
  elements and embedded tags is lost and multiple levels of mixed
  content does not work
\item  If a leaf element has both a text element and attributes then the text
  must be accessed through a vector (to provide a container for the
  attribute)
\end{itemize}
In general however this format is relatively useful.


\subsection{Options}

\begin{verbatim}
simpleTreeHandler.options.noReduce = { <tag> = bool,.. }

    - Nodes not to reduce children vector even if only 
      one child

domHandler.options.(comment|pi|dtd|decl)Node = bool 

    - Include/exclude given node types
\end{verbatim}
\subsection{Usage}

Pased as delegate in xmlParser constructor and called as callback by
xmlParser:parse(xml) method.

\section{History}

This library is fork of LuaXML library originaly created by Paul
Chakravarti. Its original version can be found at
\url{http://manoelcampos.com/files/LuaXML--0.0.0-lua5.1.tgz}. Some files not
needed for use with luatex were droped from the distribution.
Documentation was converted from original comments in the source code.

\section{License}

This code is freely distributable under the terms of the Lua license
(\url{http://www.lua.org/copyright.html})
\end{document}
